Cerebral oxygenation manuscript figures

### 2022-01-10 final_figures.Rmd
updated final figures with final analysis data

# Load libraries
library(tidyverse)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

# Load data
load("psd_plot_data.Rdata")
load("glm_model.Rdata")
load("results.Rdata")
load("roc_curves.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")
load("all_dfa_results.Rdata")

@@

  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
    as_tibble() %>% 
    dplyr::select(-1) %>% 
    mutate(Status = replace(Status, Status == "HV", "HC")) %>% 
    mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))

# pivot long

  df_hbox_long <- df_hbox %>% pivot_longer(cols = 7:ncol(.), names_to = "variable", values_to = "value")

@@

```{r table1}

# number of subjects 
df_status <- df_hbox %>% 
  count(Status) %>% 
  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  group_by(Status) %>% 
  count(Sex) %>% 
  dplyr::filter(Sex != "") %>% 
  pivot_wider(names_from = "Sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%

  pivot_wider(c(1,4), names_from = "Status", values_from = "perc_fem") %>% 
  mutate(variable = "Sex (% female)") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

@@

df_clinical_res <- df_hbox %>%

  # dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%

  group_by(Status) %>% 
  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_q1 = quantile(Age, 0.25, na.rm = TRUE), age_q3 = quantile(Age, 0.75, na.rm = TRUE),

            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_q1 = quantile(Temperature, 0.25, na.rm = TRUE), temp_q3 = quantile(Temperature, 0.75, na.rm = TRUE),
            `Heart rate` = median(HR, na.rm = TRUE), hr_q1 = quantile(HR, 0.25, na.rm = TRUE), hr_q3 = quantile(HR, 0.75, na.rm = TRUE),
            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_q1 = quantile(BP.Systolic, 0.25, na.rm = TRUE), sys_q3 = quantile(BP.Systolic, 0.75, na.rm = TRUE),
            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_q1 = quantile(BP.Diastolic, 0.25, na.rm = TRUE), dia_q3 = quantile(BP.Diastolic, 0.75, na.rm = TRUE),
            `Respiration rate` = median(RR, na.rm = TRUE), rr_q1 = quantile(RR, 0.25, na.rm = TRUE), rr_q3 = quantile(RR, 0.75, na.rm = TRUE),
            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_q1 = quantile(O2.Sat, 0.25, na.rm = TRUE), o2_q3 = quantile(O2.Sat, 0.75, na.rm = TRUE),

            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_q1 = quantile(Hematocrit, 0.25, na.rm = TRUE), hem_q3 = quantile(Hematocrit, 0.75, na.rm = TRUE),
            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lac_q1 = quantile(Lactate, 0.25, na.rm = TRUE), lac_q3 = quantile(Lactate, 0.75, na.rm = TRUE),
            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_q1 = quantile(Arginine.umol.L, 0.25, na.rm = TRUE), arg_q3 = quantile(Arginine.umol.L, 0.75, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, 1) %>%

  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_q1, ", ", age_q3, ")"),

         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_q1, ", ", temp_q3, ")"),
         `Heart rate` = paste0(`Heart rate`, " (", hr_q1, ", ", hr_q3, ")"),
         `Systolic BP` = paste0(`Systolic BP`, " (", sys_q1, ", ", sys_q3, ")"),
         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_q1, ", ", dia_q3, ")"),
         `Respiration rate` = paste0(`Respiration rate`, " (", rr_q1, ", ", rr_q3, ")"),
         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_q1, ", ", o2_q3, ")"),

         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_q1, ", ", hem_q3, ")"),
         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lac_q1, ", ", lac_q3, ")"),
         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_q1, ", ", arg_q3, ")")) %>% 
  dplyr::select(-c(age_q1, age_q3, 
                   temp_q1, temp_q3, hr_q1, hr_q3, sys_q1, sys_q3, dia_q1, dia_q3, rr_q1, rr_q3, o2_q1, o2_q3, 
                   hem_q1, hem_q3, lac_q1, lac_q3, arg_q1, arg_q3)) %>%

  unique() %>% 
  t() %>% 
  row_to_names(row_number = 1) %>% 
  as_tibble(rownames = "variable") %>% 
  add_row(df_status, .before = 1) %>% 
  add_row(df_sex, .before = 2) %>%

  dplyr::rename(" " = "variable",
                "Healthy children" = "HC",
                "Uncomplicated malaria" = "UM",
                "Cerebral malaria" = "CM")

  # remove lactate data for HC
  df_clinical_res[11,]$`Healthy children` <- "--"

  # export to CSV
  # write.csv(df_clinical_res, file = "Outputs/table_1.csv")

@@

```
**Table 1.** A summary of subject information and hematological data. Group patient data presented as median (quartile 1, quartile 3). Note: Sex not recorded for 3 CM patients; Lactate data removed for healthy volunteers due to inconsistent data.

### Fig 0. (optional) Power spectral density curves

```{r fig0}

# plot
f_plot_psd <- function(df, status, upper_lim) {

  df %>% 
    dplyr::filter(Status == status) %>% 
    dplyr::filter(freq > 0.01 & freq < upper_lim) %>% 
    ggplot(aes(x = freq, y = med_power)) +
    geom_line() +
    geom_vline(xintercept = 0.1, color = "red", lty = 2) +
    facet_grid(tissue ~ variable, scales = "free_y") +
    labs(x = "Frequency", y = "Median power across patients") +
    theme_bw()

## 0-1 Hz

  # HC
  f_plot_psd(df_psd, "HC", 1) + ggtitle("Healthy children")

  # UM
  f_plot_psd(df_psd, "UM", 1) + ggtitle("Uncomplicated malaria")

  # CM
  f_plot_psd(df_psd, "CM", 1) + ggtitle("Cerebral malaria")

## 0-3 Hz

  # HC
  f_plot_psd(df_psd, "HC", 3) + ggtitle("Healthy children")

  # UM
  f_plot_psd(df_psd, "UM", 3) + ggtitle("Uncomplicated malaria")

  # CM
  f_plot_psd(df_psd, "CM", 3) + ggtitle("Cerebral malaria")

@@

# extract variables of interest
voi <- df_all_dfa_results %>% colnames() %>% .[-(1:2)]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% 
  dplyr::filter(variable %in% voi & !grepl("fu", variable)) %>%

  mutate(tissue = as.factor(ifelse(grepl("muscle", variable), "muscle", "brain")),
         measurement = factor(ifelse(grepl("conc", variable), "Hb_tot", "Hb_oxy"), levels = c("Hb_oxy", "Hb_tot")),
         metric = case_when(

           grepl("overall_alpha|conc_alpha|o2sat_alpha", variable) ~ "alpha",
           grepl("second_alpha", variable) ~ "second_alpha",
           TRUE ~ "mean"

         )) %>% 
  dplyr::select(-variable)

@@

# plot
df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", metric))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_grid(measurement ~ tissue, scales = "free_y") +

@@

# plot
df_hbox_voi_long %>% dplyr::filter((grepl("second_alpha", metric))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_grid(measurement ~ tissue) +

  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +

  ylim(0, 1.5) +
  scale_color_manual(values = my_col) +
  labs(x = "") +
  theme_bw() +
  theme(legend.position = "none")  +

@@

### Fig 3. Logistic regression modeling of uncomplicated and cerebral malaria subjects

@@

# include variables of interest

df_hbox_log_reg <- df_hbox %>% 
  dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))

@@

  df_hbox_impute <- df_hbox_log_reg %>% 
    group_by(Status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

@@

f_calc_cor <- function(df, status) {

  m <- df %>%
    dplyr::filter(Status == status) %>% 
    dplyr::select(-c(subject_id, Status)) %>% 
    cor(method = "spearman")

@@

# create run correlations of potential regressors
m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%

@@

  as_ggplot(p_leg)

@@

```{r fig3b}
# remove healthy volunteers for logistic regression analysis
df_hbox_impute <- df_hbox_impute %>% dplyr::filter(Status != "HC")

# convert tibble to df
df_hbox_df <- df_hbox_impute %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  as.data.frame()

# add rownames
rownames(df_hbox_df) <- df_hbox_impute %>% .$subject_id

# compute PCA
my_pca <- prcomp(df_hbox_df, scale = TRUE)

# define groups
groups <- as.factor(df_hbox_impute$Status)

@@

# two data frames for plotting
df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)

@@

  # plot patients
  ggplot() +
  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), size = 3) +
  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  # geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
  # geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +

@@

#### C. Coefficients
```{r table2c}

@@

table2c <- hbox_final %>%
  tidy(exponentiate = TRUE) %>%
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  as.data.frame()

table2c %>% knitr::kable(align = rep('c', 5))

@@

```{r fig3d_prep, eval = FALSE}

## write function to create model on given df

f_log_reg_model <- function(df_model) {

  # resample
  hbox_boot <- bootstraps(df_model, strata = Status, times = 1000)

  # recipe
  hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
    step_BoxCox(all_predictors()) %>% 
    step_nzv(all_predictors()) %>% 
    step_smote(Status) # upsample to deal with unbalanced sample sizes

  # create model on resamples
  glm_res <- hbox_wf %>% 
    add_model(glm_spec) %>%
    fit_resamples(
      resamples = hbox_boot,
      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
    )

  # finalize model
  hbox_final <- hbox_wf %>% 
    add_model(glm_spec) %>% 
    finalize_workflow(glm_best) %>%
    fit(df_model) %>%
    pull_workflow_fit()

  # view coefficients & p-values
  model_coefs <- hbox_final %>% tidy()

  # return list including model and a table of best model variable statistics
  list(glm_res, model_coefs)

## write function to collect predictions for ROC plot

f_collect_preds <- function(glm_res) {

      glm_res %>% 
      collect_predictions() %>% 
      roc_curve(Status, .pred_CM)

## Perform functions on all iterations of model

set.seed(123)

    # select starting variables
    df_hbox_model1 <- df_hbox_log_reg %>% 
      dplyr::select(c(Status, Hematocrit, Lactate, avg_Hb_o2sat, Hb_o2sat_second_alpha, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
      mutate(Status = factor(Status, levels = c("CM", "UM")))

    # model 1: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + Hb_o2sat_second_alpha + avg_Hb_conc + Hb_conc_second_alpha
    model1 <- f_log_reg_model(df_hbox_model1)
    roc1 <- f_collect_preds(model1[[1]]) %>% mutate(model = "1")
    model1[[2]] # remove Hb_o2sat_second_alpha (highest p-val)

    # model 2: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
    df_hbox_model2 <-  df_hbox_model1 %>% dplyr::select(-Hb_o2sat_second_alpha)
    model2 <- f_log_reg_model(df_hbox_model2)
    roc2 <- f_collect_preds(model2[[1]]) %>% mutate(model = "2")
    model2[[2]] # remove Lactate

    # model 3: Status ~ Hematocrit + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
    df_hbox_model3 <-  df_hbox_model2 %>% dplyr::select(-Lactate)
    model3 <- f_log_reg_model(df_hbox_model3)
    roc3 <- f_collect_preds(model3[[1]]) %>% mutate(model = "3")
    model3[[2]] # remove avg_Hb_o2sat

    # model 4: Status ~ Hematocrit + avg_Hb_conc + Hb_conc_second_alpha
    df_hbox_model4 <-  df_hbox_model3 %>% dplyr::select(-avg_Hb_o2sat)
    model4 <- f_log_reg_model(df_hbox_model4)
    roc4 <- f_collect_preds(model4[[1]]) %>% mutate(model = "4")
    model4[[2]] # none to remove, final model

save(roc1, roc2, roc3, file = "roc_curves.Rdata")

# calculate areas under curves

library(MESS)
roc1 <- roc1 %>% mutate(false_pos = 1 - specificity)
auc(roc1$false_pos, roc1$sensitivity)

roc2 <- roc2 %>% mutate(false_pos = 1 - specificity)
auc(roc2$false_pos, roc2$sensitivity)

roc3 <- roc3 %>% mutate(false_pos = 1 - specificity)
auc(roc3$false_pos, roc3$sensitivity)

roc_final <- glm_res %>% 
  collect_predictions() %>% 
  roc_curve(Status, .pred_0) %>% 
  mutate(false_pos = 1 - specificity)
auc(roc_final$false_pos, roc_final$sensitivity)

@@

df_roc <- glm_res %>% 
  collect_predictions() %>% 
  roc_curve(Status, .pred_0) %>% 
  mutate(model = "final") %>% 
  bind_rows(roc1, roc2, roc3)

ggplot() +
  geom_path(data = df_roc %>% filter(model == "final"), 
            aes(x = 1 - specificity, y = sensitivity, color = model), 
            size = 2.5) +
  geom_abline(lty = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  theme_bw() +

@@

fig3d_inset <- collect_metrics(glm_res) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

### Fig 4. Explanatory model of how to interpret hemoglobin concentration

@@

#### A. Pre-processed signal
```{r fig_s3a}

df_thc_cumsum1 %>%

  ggplot(aes(x = Time, y = sg_filt)) +
  geom_line() +

>>>>

# Load libraries
library(tidyverse)
library(readxl)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

# Load data
load("objects/glm_model.Rdata")
load("objects/bandpass_filtered_dfa_results.Rdata")
load("objects/sims.Rdata")

@@

df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status)

@@

#### Table 1a. All clinical data
```{r table1a, eval = FALSE}

# number of subjects 
df_status <- df_hbox %>% 
  count() %>% 
  pivot_wider(1:2, names_from = "status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  count(sex) %>% 
  dplyr::filter(sex != "") %>% 
  pivot_wider(names_from = "sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%

  pivot_wider(c(1,4), names_from = "status", values_from = "perc_fem") %>% 
  mutate(variable = "Sex (% female)") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

@@

  # select variables for table
  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")

  # calculate median, first and third quartiles
  df_clinical_res <- df_hbox %>% 
    select_at(clinical_voi) %>% 
    summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
    mutate_if(is.numeric, round, 2)

  # write function to paste together values
  f_unite_cols <- function(col) {
    df_clinical_res %>% 
      unite(!! col, starts_with(col), sep = "_") %>% 
      dplyr::select(!! col) 
    }

  # paste together and transpose
  df_clinical_res <- 1:length(clinical_voi) %>% 
    purrr::map_dfc(~f_unite_cols(clinical_voi[.x])) %>% 
    mutate_all(funs(str_replace(., "_", " ("))) %>% 
    mutate_all(funs(str_replace(., "_", ", "))) %>% 
    mutate_all(funs(paste0(., ")"))) %>%

    t() %>% 
    as_tibble(rownames = "variable") %>% 
    dplyr::rename("HC" = "V1", "UM" = "V2", "CM" = "V3") %>% 
    add_row(df_status, .before = 1) %>% 
    add_row(df_sex, .before = 2)

# export to CSV
  write.csv(df_clinical_res, file = "Outputs/table_1.csv")

@@

```
**Table 1a.** A summary of subject information and hematological data. Group patient data presented as median (quartile 1, quartile 3).

#### Table 1b. CM-specific data
```{r table1b, eval = FALSE}

cm_voi <- c("bcs_session", "brain_swell", "hrp2")

df_cm_voi <- df_hbox %>% 
  ungroup %>% 
  filter(status == "CM") %>% 
  select_at(cm_voi) %>% 
  mutate_at(c("bcs_session", "brain_swell"), as.factor)

# BCS counts
df_cm_voi %>% count(bcs_session) %>% knitr::kable()

# Brain swell score counts
df_cm_voi %>% count(brain_swell) %>% knitr::kable()

# HRP
df_cm_hrp2 <- df_cm_voi %>% 
  select(cm_voi[3]) %>% 
  summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
  unite(hrp2, 1:2, sep = " (") %>% 
  unite(hrp2, 1:2, sep = ", ") %>% 
  mutate(hrp2 = paste0(hrp2, ")"))

df_cm_hrp2 %>% knitr::kable()

```
**Table 1b.** CM-specific clinical characteristics

@@

# extract hb_tot and hb_oxy for the brain and muscle
df_fig1 <- df_hbox %>% 
  select(contains(c("hb_tot","hb_oxy")), -contains(c("alpha", "bsl"))) %>%
  ungroup() %>% 
  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge")

@@

# plot
df_fig1 %>%

  ggplot(aes(x = status, y = value)) +
  geom_violin(aes(color = status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
  facet_grid(variable ~ tissue, scales = "free_y") +

@@

# extract hb_tot and hb_oxy for the brain and muscle
df_fig2 <- df_hbox %>% 
  select(contains("alpha2")) %>%
  ungroup() %>% 
  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge") %>% 
  mutate(variable = str_remove(variable, "2"))

# plot
df_fig2 %>%

  ggplot(aes(x = status, y = value)) +
  geom_violin(aes(color = status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
  facet_grid(variable ~ tissue, scales = "free_y") +

  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +

  scale_color_manual(values = my_col) +
  ylim(c(0, 1.5)) +
  labs(x = "") +
  theme_bw() +
  theme(legend.position = "none")  +

@@

### Fig 3. Increased cerebral_hb_tot and decreased cerebral_hb_tot_alpha are distinct clinical features of cerebral malaria

@@

# select model variables of interest
model_voi <- c("subject_id", "status", 
               "hct", "lactate",
               "cerebral_hb_tot", "cerebral_hb_oxy",
               "cerebral_hb_tot_alpha2", "cerebral_hb_oxy_alpha2")

df_model <- df_hbox %>% 
  select_at(model_voi) %>%

  # remove HC
  ungroup %>% 
  filter(status != "HC")

@@

  df_model_impute <- df_model %>% 
    group_by(status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

@@

f_calc_cor <- function(df, group) {

  m <- df %>%
    dplyr::filter(status == group) %>% 
    dplyr::select(-c(subject_id, status)) %>% 
    cor(method = "spearman")

@@

# create run correlations of potential regressors
m_cor_um <- df_model_impute %>% f_calc_cor(group = "UM")

m_cor_cm <- df_model_impute %>% f_calc_cor(group = "CM")

@@

    mutate_if(is.factor, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%

@@

  # as_ggplot(p_leg)

@@

```{r fig3b}

# convert tibble to df
df_model_pca <- df_model_impute %>% 
  dplyr::select(-c(subject_id, status)) %>% 
  as.data.frame()

# add rownames
rownames(df_model_pca) <- df_model_impute$subject_id

# compute PCA
my_pca <- prcomp(df_model_pca, scale = TRUE)

# define groups
groups <- as.factor(df_model_impute$status)

@@

# two data frames for plotting
df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
df_pnts <- my_pca$x %>% 
  as_tibble() %>% 
  mutate(status = groups, 
         subject_id = df_model_impute$subject_id)

@@

  # plot patients
ggplot() +
  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status), size = 3) +
  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, fill = status),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  # stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, shape = status),
  #           na.rm = FALSE) +

  # View patient IDs
  # geom_text_repel(data = df_pnts, aes(x = PC1, y = PC2, label = subject_id, size = 10), show.legend = FALSE) +

  # plot variables
  # geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
  geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +

@@

#### C. Coefficients

```{r model, eval = FALSE}

### shows how logistic regression model was set up and run, results saved in objects/glm_model ###

# resample
set.seed(234)
hbox_boot <- bootstraps(df_model, strata = Status, times = 5000)

# recipe
hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
  step_impute_knn(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_smote(Status) %>% 
  update_role(subject_id, new_role = "id")

# create model on 5000 different resamples
glm_res <- hbox_wf %>% 
  add_model(glm_spec) %>%
  fit_resamples(
    resamples = hbox_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# finalize model
hbox_final <- hbox_wf %>% 
  add_model(glm_spec) %>% 
  finalize_workflow(glm_best) %>%
  fit(df_model) %>%
  extract_fit_parsnip()

```{r table3c}

@@

table3c <- hbox_final %>%
  tidy(exponentiate = TRUE) %>%
  filter(p.value < 0.05) %>% 
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  as.data.frame()

table3c %>% knitr::kable(align = rep('c', 5))

@@

@@

df_roc <- glm_res %>% 
  collect_predictions() %>% 
  roc_curve(status, .pred_CM)

ggplot() +
  geom_path(data = df_roc, 
            aes(x = 1 - specificity, y = sensitivity), 
            size = 2.5) +
  geom_abline(lty = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  theme_bw() +

@@

fig3d_inset <- glm_res %>% 
  collect_metrics() %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

### Fig 4. An increase in cerebral_hb_tot is significantly associated with an increase in brain swelling score

```{r fig4_plot}

# select model voi
brain_swell_voi <- c("brain_swell", "cerebral_hb_tot", "hct")
df_fig4 <- df_hbox %>% select_at(brain_swell_voi)

# remove cerebral_hb_tot outliers
  df_fig4_outliers <- df_fig4 %>% 
    mutate(Q1_hbtot = quantile(cerebral_hb_tot, 0.25, na.rm = TRUE),
           Q3_hbtot = quantile(cerebral_hb_tot, 0.75, na.rm = TRUE),
           IQR_hbtot = Q3_hbtot - Q1_hbtot,
           outlier_hbtot = case_when(

             cerebral_hb_tot < Q1_hbtot - IQR_hbtot ~ 1,
             cerebral_hb_tot > Q3_hbtot + IQR_hbtot ~ 1,
             TRUE ~ 0

           )

    ) %>% 
    filter(outlier_hbtot != 1)

# model
  lm_fig4 <- lm(brain_swell ~ cerebral_hb_tot + hct, data = df_fig4_outliers)
  f <- summary(lm_fig4)$fstatistic
  p <- pf(f[1], f[2], f[3], lower.tail = FALSE)

  model <- "model: brain_swell ~ cerebral_hb_tot + hematocrit"
  r2 <- paste0("adjusted R^2: ", lm_fig4 %>% summary() %>% .$adj.r.squared %>% round(2))
  model_p_val <- paste0("model: ", round(p, 3))
  hb_tot_p_val <- paste0("cerebral_hb_tot: ", lm_fig4 %>% tidy() %>% .[2, "p.value"] %>% .$p.value %>% round(3))

# plot variables
require(ggpmisc)

df_fig4_outliers %>% 
  ggplot(aes(x = cerebral_hb_tot, y = brain_swell)) +
  geom_point(shape = 1, aes(size = hct, color = as.factor(brain_swell))) +
  #geom_abline(intercept = ) +

  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2) +
  annotate("text", label = model, x = 25, y = 8, hjust = 0, vjust = 1, size = 5) +
  annotate("text", label = r2, x = 25, y = 7.5, hjust = 0, vjust = 1, size = 5) +
  annotate("text", label = "p-values", x = 25, y = 7, hjust = 0, vjust = 1, size = 5) +
  annotate("text", label = model_p_val, x = 35, y = 6.75, hjust = 0, vjust = 1, size = 4) +
  annotate("text", label = hb_tot_p_val, x = 35, y = 6.5, hjust = 0, vjust = 1, size = 4) +

  lims(x = c(25, 225)) +
  theme_bw() +

  theme(# text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# 3d plot?  
# require(plotly)
#   x <- df_fig4_outliers$cerebral_hb_tot
#   y <- df_fig4_outliers$hematocrit
#   z <- df_fig4_outliers$brain_swell
#   
#   df_fig4_outliers %>% plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", color = x)

### Fig 5. Explanatory model of how to interpret hemoglobin concentration

@@

#### A. Raw signal
```{r fig_s1a}

df_figs1a <- read.table("Data/signal_segments/Hb_tot/TM0001CM01.txt")

df_figs1a %>%

  ggplot(aes(x = Time)) +
  geom_line(aes(y = THC)) +

  labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
  ggtitle("Raw signal") +
  theme_bw() +
  theme(# text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))
